Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-07-05 submission/manuscript.Rmd
editing intro

Embryonic stem cells employ a single genome to form a myriad of discrete cellular identities. This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through histone and DNA modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001]. Although many chromatin regulators were initially identified for their roles in cellular identity [@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a], recent advancements in next generation sequencing unexpectedly revealed many neurodevelopmental disorders (NDDs) are caused by or linked to mutations in chromatin regulators[CIT]. The connection between neurodevelopment and chromatin regulation can be partially explained by their regulation of brain-specific genes or chromatin states, such as orchestrating synaptic maturation[@zhouBrainspecificPhosphorylationMeCP22006] or the transition between neuron and glia during cortical development[@hirabayashiPolycombLimitsNeurogenic2009]. However, loss of some chromatin regulators can also result in aberrant transcription of tissue-specific genes outside of their intended environment[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017]. Very few studies have explored the ectopic expression of tissue-specific genes in chromatin-linked neurodevelopmental disorders[@schaeferControlCognitionAdaptive2009; @scandagliaLossKdm5cCauses2017] and it is unclear if this partial loss of brain identity contributes to neurodevelopmental impairments.

To elucidate the contribution of tissue identity dysregulation to neurodevelopmental impairments, it is essential to first characterize the types of genes misexpressed, the developmental time point the dysregulation begins, and the molecular mechanism underlying their de-repression. Characterizing these features will enable us to identify molecular footholds common between NDDs that can then be exploited for potential therapeutics.

In this study, we characterized the dysregulation of tissue-enriched genes with loss of the chromatin regulator lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C), a histone 3 lysine 4 demethylase. Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534), whose features include short stature, intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020]. Previous work has demonstrated constitutive _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ patient phenotypes, including hyperaggression and learning impairments[@iwaseMouseModelXlinked2016]. Next generation RNA sequencing (RNA-seq) in the _Kdm5c_-KO hippocampus unexpectedly revealed ectopic expression of testis-enriched genes within the brain[@scandagliaLossKdm5cCauses2017]. However, it is currently unclear if misexpression in the _Kdm5c_-KO brain is unique to testis genes, as other tissue-enriched genes have not been systematically evaluated.

To elucidate KDM5C's role in tissue identity, here we characterized the aberrant transcription of tissue-enriched genes within the _Kdm5c_-KO brain and epiblast-like cells (EpiLCs), an _in vitro_ model of the post-implantation embryo. We identified general dysregulation of tissue-enriched genes in both the adult _Kdm5c_-KO brain and EpiLCs, including misexpression of liver, muscle, and ovary genes. The _Kdm5c_-KO amygdala and hippocampus had significant enrichment of testis-biased genes that are unique to germ cells. To better characterize germline gene misexpression, we then generated a dataset of germline-enriched genes by comparing gene expression in gonads with germ cell depletion. We found _Kdm5c_-KO EpiLCs primarily expressed unique germline genes compared to the mature _Kdm5c_-KO brain, including _Dazl_ and _Stra8_, key drivers of germline identity and meiosis. While KDM5C is directly bound to some germline gene promoters in EpiLCs, it is not directly bound to many germline-enriched mRNAs expressed with _Kdm5c_-KO cells, indicating germline genes can be aberrantly transcribed through indirect mechanisms. Finally, we found KDM5C loss impairs the placement of DNA methylation at germline gene promoters as ESCs differentiate into EpiLCs. Therefore, we propose KDM5C plays a crucial role in the development of tissue identity during early embryogenesis, including establishment of the soma-germline boundary.

**note: need a better conclusion sentence - work on when we know what's happening with last figure/functional consequences**

@@

>>>>

Embryonic stem cells employ a single genome to form the myriad of discrete cellular identities that make up a complete organism. This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through histone and DNA modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001]. Although many chromatin regulators were initially identified for their roles in cellular identity [@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a], recent advancements in next generation sequencing unexpectedly revealed many neurodevelopmental disorders (NDDs) are caused by or linked to mutations in chromatin regulators[CIT]. The connection between neurodevelopment and chromatin regulation can be partially explained by their regulation of brain-specific genes or chromatin states, such as orchestrating synaptic maturation[@zhouBrainspecificPhosphorylationMeCP22006] or the transition between neuron and glia during cortical development[@hirabayashiPolycombLimitsNeurogenic2009]. However, loss of some chromatin regulators can also result in aberrant transcription of tissue-specific genes outside of their intended environment[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017]. Very few studies have explored the ectopic expression of tissue-specific genes in chromatin-linked neurodevelopmental disorders[@schaeferControlCognitionAdaptive2009; @scandagliaLossKdm5cCauses2017] and it is unclear if this partial loss of brain identity contributes to neurodevelopmental impairments.

To elucidate if dysregulation of brain tissue identity contributes to neurodevelopmental impairments, it is essential to first characterize the types of genes misexpressed and the molecular mechanisms underlying their de-repression. In this study, we characterized the dysregulation of tissue-enriched genes with loss of the chromatin regulator lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C), a histone 3 lysine 4 demethylase. Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534), whose features include short stature, intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020]. Previous work has demonstrated constitutive _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ patient phenotypes, including hyperaggression and learning impairments[@iwaseMouseModelXlinked2016]. Next generation RNA sequencing (RNA-seq) in the _Kdm5c_-KO hippocampus unexpectedly revealed ectopic expression of testis-enriched genes within the brain[@scandagliaLossKdm5cCauses2017]. However, it is currently unclear if misexpression in the _Kdm5c_-KO brain is unique to testis genes, as other tissue-enriched genes have not been characterized. Furthermore, some of the ectopic testis transcripts have known functions in germ cells[@scandagliaLossKdm5cCauses2017], potentially reflecting a failure to distinguish between somatic (non-meiotic) and germline (meiotic) identities with KDM5C loss. While there is emerging evidence that KDM5C regulates select germline genes, it is unclear which germline  _Kdm5c_-KO.
- KDM5C's molecular mechanisms unclear - direct vs indirect
- Developmental time point of involvement
- How the genes change over time.

- Systematically characterizing KDM5C's role in repressing germline identity will provide molecular footholds for testing the impact of ectopic germline gene expression on neurodevelopment. These can then be developed into novel therapeutics for MRXSCJ treatment and other NDDs that share a similar etiology
**note: don't want to be too specific about therapeutics**

To elucidate KDM5C's role in tissue identity, here we characterized the aberrant transcription of tissue-enriched genes within the _Kdm5c_-KO brain and epiblast-like cells (EpiLCs), an _in vitro_ model of the post-implantation embryo. We identified general dysregulation of tissue-enriched genes in both the adult _Kdm5c_-KO brain and EpiLCs, including misexpression of liver, muscle, ovary, and testis genes. Both the _Kdm5c_-KO amygdala and hippocampus had significant enrichment of testis-biased genes that are typically only expressed in germ cells. While the _Kdm5c_-KO brain primarily expresses genes important for late germ cell development, _Kdm5c_-KO EpiLCs aberrantly express key drivers of germline identity and meiosis, such as _Dazl_ and _Stra8_. We found that KDM5C only binds a subset of germline genes expressed in the _Kdm5c_-KO brain and EpiLCs, indicating germline-enriched mRNAs can be aberrantly expressed through indirect mechanisms. Finally, we found KDM5C loss impairs the placement of DNA methylation at germline gene promoters as ESCs differentiate into EpiLCs. Therefore, we propose KDM5C plays a crucial role in the development of tissue identity during early embryogenesis, including establishment of the soma-germline boundary.

- We then generated a dataset of germline-enriched genes to aid in the characterization of germline gene misexpression
- **note: need a better conclusion sentence - work on when we know what's happening with last figure/functional consequences**
- **note: I feel like I'm missing the key points still on the last paragraph**

@@

**Germline gene repression background:**
